Machine Learning guided workflow for Ribosome Binding Site engineering

### 2021-05-06 paper/main.tex
Updated main text.

This way, were able to build an extensive, reliable library of novel RBSs with diverse sequences.

@@

Figure \ref{fig: Flowchart} shows the DBTL cycle with the machine learning process emphasised. Our machine learning workflow encompasses two phases of the cycle, namely the LEARN and DESIGN phase. 
In LEARN phase we use the Gaussian Process regression algorithm to predict the TIR of different RBS sequences and in the DESIGN phase we use the Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be Built and Tested next. 
The following sections will discuss our machine learning workflow divided into specific parts of the DBTL cycle.\\

@@

\subsection{DESIGN of the genetic device}
There is a number of factors that impact the protein expression rate, many of them concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992,SHULTZABERGER2001}.
In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. 
By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring plasmid with the device over time.
Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same time with addition of IPTG.\\
In \emph{E. coli} the RBS is usually located in the 20 bases upstream of the start codon. 
Additionally, there is a consensus RBS core sequence called the Shine-Dalgarno sequence, which in \emph{E. coli} is AGGAGG. 
Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure \ref{fig:Anatomy}).\\

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.6]{plots/RBS_anatomy.pdf}
    \caption{\textbf{Diagram of the investigated sequence.} The sequence of the investigated RBS is shown with the randomised core sequence in red. The start codon of the GFP coding sequence is also shown with numbers showing relative nucleotide positions.}
    \label{fig:Anatomy}
\end{figure}

The investigated RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACAT, which is a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}. 
In our design we focus on randomising of the core -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the consensus sequence, i.e. TTTAAGA + NNNNNN + TATACAT.
Since for each of the 6 position there are 4 possibilities: A, C, G, T the total experimental (variant) space is $4^6$ = 4096.\\
To generate the dataset that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, which constitutes a little over 10\% of the whole experimental space. 
To fit into our automated workflow, we have divided the 450 variants into batches of 90.
In the zeroth round we have tested two batches of designs, for total of 180 variants. 
Since at that stage we didn't have access to any prior data fitting our design we approximated the predictions using the data from \textcite{jervis2018machine} and the Gaussian Process regression method (see \textbf{Methods} section) \cite{srinivas2012information}.
This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
This dataset has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.\\
Apart from the abovementioned algorithm recommnded sequences, we have also decided to try some random sequences for the total of 180 sequences:

\begin{enumerate}
    \item 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence. This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part.
    \item 30 RBS sequences that were fully and uniformly randomised (equal probability of choosing either nucleotide for each position) 
    \item 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occuring RBS sequences in \emph{E. coli} genome
    \item 60 RBS sequences recommended by our multi armed bandit recommendation algorithm based on the initial literature, as described above.
\end{enumerate}{}

In the subsequent rounds, data obtained in the previous ones have been used to teach the algorithm. 
These subsequent rounds were run in batches of 90 each. \\

@@

For the machine learning to work effectively the analysed data-set needs to show two qualities: high relative volume and high quality of data.
These two elements don't have a specific definitions, but in general the data-set have to be going into at least hundreds possible data points and one have to cover 5-10\% of that space.
And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
This way we hope to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down time to prepare our variants.\\
In short, the genetic variations of the RBS were introduced to the plasmids with combination of PCR and isothermal assembly. 
The plasmids were then transformed and the resulting transformants were tested using microplate reader.
Vast majority of reactions were prepared using liquid handling equipment.
Similarly, colony picking was done by automated colony picker.\\

@@

The three main considerations when designing a machine learning workflow for synthetic biology is feature representation, prediction and recommendation algorithms.
Since machine learning is just starting to be used in synthetic biology, there is no agreed approach to any of these problems.\\
In our case, the problem of feature representation was how to embed the DNA sequence to a numerical vector which can be used in machine learning predictions. 
Kernel functions implicitly embed DNA sequences into high-dimensional feature space and capturing similarities between data points.  
%Gaussian Process regression has been used, we have investigated use of different types of kernels (also called covariances) for embedding \cite{Ben-Hur2008}. 
% The additional, and positive, effect of using kernels is that our data will be moved to a higher-dimensional space, which makes the regression process easier. 
% We compared performance of Dot Product, RBF and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure Xa}. 
% Since we found that the spectrum kernel performed the best, we have used it in subsequent studies.
% TODO: citation needed. 
In our case, we consider string kernels \cite{}, which takes two DNA sequences as input and output a real value which represents the similarity between those two sequences. 
The commonly used string kernels includes spectrum kernels \cite{}, mixed spectrum kernels \cite{}, weighted degree kernels \cite{} and weighted degree kernels with shifts \cite{}. 
We review the design of the above string kernel in Appendix \ref{}.
For Round 0, we chose to use the spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.
For other rounds, we chose to use the weighted degree kernel with shifts. \\
% TODO: reasoning needed. 
% This approach allowed for good balance between computational complexity and performance.\\

@@

The recommendations were made using the Multi-armed Bandit algorithm.
In short, the this algorithm is a stochastic method of probing of the experimental space. 
This algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital). 
In our case we use the Upper Confidence Bound version of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computed by the prediction algorithm (GPR). 
Another feature of the bandits algorithms is that it balances two approaches: exploration and exploitation. 
Exploration makes the algorithm recommend designs that will improve the predictions, whereas exploitation will recommend designs that focus on delivering the most efficient design the fastest. 
The two approaches can be controlled with the \textbeta\enspace parameter. 
We have decided that in the first iterations of the cycle it would be beneficial to skew the algorithm towards the exploration with exploitation taking increasing role in later iterations. 
One thing of note is that the bandit algorithm is stochastic, that is it exploits the probabilities of given event occurring (in this case RBS having a specific TIR). 
As such, it pairs naturally with our prediction algorithm, the Gaussian Process, which provides probability based function regression.
\\

>>>>

This way, we were able to build an extensive, reliable library of novel RBSs with diverse sequences.

@@

The three main considerations when designing a machine learning workflow for synthetic biology is feature representation, prediction and recommendation algorithms.
We will discuss our workflow in terms of the Design-Build-Learn-Test cycle, where prediction will be discussed in the Design section and feature representation with prediction in the Learn one.\\
Figure \ref{fig: Flowchart} shows the DBTL cycle with the machine learning process emphasised.
Principally, machine learning is used in the LEARN and DESIGN phase and the other two are driven by choices made by human researchers. 
In LEARN phase we use the Gaussian Process regression algorithm to predict the TIR of different RBS sequences and in the DESIGN phase we use the Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be Built and Tested next.\\

@@

\subsection{DESIGN of the genetic device}
In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. 
By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring plasmid with the device over time.
Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same time with addition of IPTG.\\

In \emph{E. coli} the RBS is usually located in the 20 bases upstream of the start codon. 
Additionally, there is a consensus RBS core sequence called the Shine-Dalgarno sequence, which in \emph{E. coli} is AGGAGG. 
Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (see detail in Figure \ref{fig:Flowchart}).\\

Our template RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACA.
This sequence is a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}. 
Since this is the sequence against which new RBS sequenced will be benchmarked we will refer to this sequence as the benchmark sequence from now on.
In our design we focus on randomising of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the consensus sequence, i.e. TTTAAGA + NNNNNN + TATACAT.
We have experimentally confirmed that changing the core sequence is statistically more impactful on TIR than changes made outside of it (see Supplementary materials). 
Since for each of the 6 position there are 4 possibilities: A, C, G, T the total experimental (variant) space is $4^6$ = 4096.\\

The design recommendations were made using the Multi-armed Bandit algorithm.
In short, the this algorithm is a stochastic method of probing of the experimental space. 
This algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital). 
In our case we use the Upper Confidence Bound version of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computed by the prediction algorithm (GPR). 
Another feature of the bandits algorithms is that it balances two approaches: exploration and exploitation. 
Exploration makes the algorithm recommend designs that will improve the predictions, whereas exploitation will recommend designs that focus on delivering the most efficient design the fastest. 
The two approaches can be controlled with the \textbeta\enspace parameter. 
We have decided that in the first iterations of the cycle it would be beneficial to skew the algorithm towards the exploration with exploitation taking increasing role in later iterations. 
One thing of note is that the bandit algorithm is stochastic, that is it exploits the probabilities of given event occurring (in this case RBS having a specific TIR). 
As such, it pairs naturally with our prediction algorithm, the Gaussian Process, which provides probability based function regression.
\\

To generate the dataset that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, which constitutes a little over 10\% of the whole experimental space. 
To fit into our automated workflow, we have divided the 450 variants into batches of 90.
In the zeroth round we have tested two batches of designs, for total of 180 variants split as below:

\begin{enumerate}
    \item 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence. This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part.
    \item 30 RBS sequences that were fully and uniformly randomised (equal probability of choosing either nucleotide for each position) 
    \item 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E. coli} genome
    \item 60 RBS sequences recommended by our implementation of recommendation algorithm based on a data set obtained from literature, as described below.
\end{enumerate}{}

To generate algorithm driven designs for the 0th round we have used the data set available in \textcite{jervis2018machine}.
This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
This data set has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.\\
In the subsequent rounds, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds. \\

@@

For the machine learning to work effectively the analysed data set needs to show two qualities: high relative volume and high quality of data.
These two elements don't have specific definitions, but in general the data set has to be going into at least hundreds data points and have to cover 5-10\% of that space.
And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
This way we hope to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down time to prepare our variants.\\
In short, the genetic variations of the RBS were introduced to the plasmids with combination of PCR and isothermal assembly. 
The plasmids were then transformed and the resulting transformants were tested using microplate reader.
Vast majority of reactions were prepared using liquid handling equipment.
Similarly, colony picking was done by an automated colony picker.\\

@@

For our predictions, we had to solve the problem of feature representation first.
In our case, that problem was how to embed the DNA sequence to a numerical vector which can be used in machine learning predictions.\\
Due to our use of the Gaussian Process Regression for predictions we have chosen kernel (covariance) functions for our embedding.
Kernel functions implicitly embed DNA sequences into high-dimensional feature space which enables capturing similarities and distances between data points.  
The additional effect of using kernels is that our data will be moved to a higher-dimensional space, which makes the regression process easier.\\

There is a number of useful kernel functions that can be used for embedding of DNA.
In our case, we use string kernels \cite{}, which take two DNA sequences as input and output a real value which represents the similarity between them. 
The commonly used string kernels includes spectrum kernels \cite{}, mixed spectrum kernels \cite{}, weighted degree kernels \cite{} and weighted degree kernels with shifts \cite{}. 
We review the design of the above string kernels in Supplementary materials \ref{}.
For Round 0, we chose to use the spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.
For subsequent rounds, we chose to use the weighted degree kernel with shifts. \\
% TODO: reasoning needed. 
% This approach allowed for good balance between computational complexity and performance.\\

@@
